### 2013-04-19 article.tex
Proof read manuscript
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. In contrast there remains a
steep learning curve in the subsequent computational steps required to process
raw sequence reads into a database-deposited genome sequence. Here we describe
``Genomer,'' a tool that simplifies the manual tasks of finishing and uploading
a genome sequence to a database. Genomer can format a genome scaffold into the
common files required for submission to GenBank. Genomer simplifies editing a
genome scaffold by allowing a human-readable YAML format file to be edited
instead of large sequence files. Genomer is written as a command line tool and
is an effort to make the manual, and increasingly common, process of genome
scaffolding more robust and reproducible. Extensive documentation and video
tutorials are available at http://next.gs.

@@

Manually editing a genome sequence or annotation is a non-trivial problem
requiring significant effort on a researcher's part. The large size of FASTA
sequence files or general feature format (GFF) annotation files makes changing
these in a text editor difficult. Tasks such as organising project files,
adding additional sequences to an assembly, updating incorrect annotations or
preparing files for database deposition all divert effort from the main goal of
analysing a genome. As a result genome projects have a steep learning curve for
researchers embarking on genome sequencing for the first time, while still
requiring effort for experienced researchers.

@@

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer'', to ease and simplify many of the
common steps in our genome project. Several graphical user interface tools are
available to work with genome projects \cite{tanaka2006, wilkinson2002,
lopez2011, carver2012, gordon2003}. Genomer instead provides a simple, command
line interface for genome finishing allowing scriptable automation and
reproducible results. We have open-sourced this tool under the MIT license and
here we describe its application for simplifying steps in a microbial genome
project.

@@

  \item[Simple editing of a draft genome sequence.]{Genomer is built on the
  existing Scaffolder \cite{barton2012} file format for assembling draft genome
  sequences. This YAML format simplifies the process of re-organising and
  trimming contigs in draft genome without having to edit large nucleotide
  sequences manually.}

  \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
  coordinates of contig annotations to respective positions in the draft genome
  sequence. This allows the scaffold file to be continuously improved and
  updated even after annotation. For instance, additional PCR sequences can be
  used to close gaps in the assembly without requiring reannotation or manually
  updating annotation coordinates.}

  \item[Generation of files for submission to GenBank.]{Submitting a genome
  sequence to GenBank database requires generating specific files. These may
  include a FASTA file of the draft sequence, a table of annotations, a FASTA
  file containing the individual contig sequences, and `a golden path' (AGP)
  \cite{agp-spec} file describing the placement of contigs. Genomer automates
  the generation of all these files from the assembled draft genome.}

@@

As Genomer is written the command-line to use in shells such as bash or zsh.
Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. This is an advantage of command-line tools
versus GUI-based tools, the latter are harder to automate and share steps taken
to produce results. Therefore Genomer may be of particular interest to
bioinformaticians who prefer automated approaches and scripting on the command
line.

Additionally as Genomer uses plain text it is easy to store versions the
project with revision control such as git \cite{git-scm}. This allows tracking
changes to the project and to reverting errors back to earlier versions.
Example build files from our genome projects can be found on GitHub for a
simple plasmid \cite{plasmid-github} and a more complex genome
\cite{genome-github} sequence.

We have released Genomer as open source on GitHub \cite{genomer-github} and
have prepared documentation and video tutorials at http://next.gs. As we have
used this tool extensively to simplify our own genome projects we believe
genomer may also be useful to others in the field. We will illustrate the
possible usefulness of genomer with two use cases: updating the annotations of
a genome scaffold during the finishing process, and preparing genome files for
submission to GenBank.

\subsection*{Example use case with Genomer}

Online tools such as the Integrated Microbial Genomes system
\cite{markowitz2006} provide an online service whereby a nucleotide sequence
may be submitted and a subsequent file of predicted genomic features returned.
Once annotated modifying the sequence becomes problematic as this requires
corresponding changes in all downstream annotated coordinates. This therefore
presents a problem whereby either no more updates can be made to the sequence
or the delta difference in annotation coordinates must be propagated to all
downstream locations.

During our own \emph{P. fluorescens} genome sequencing projects, we aimed to
continuously improve the quality of the genome sequence by closing gaps using
PCR and Sanger sequencing while at the same time analysing the observed genomic
features. We therefore incorporated functionality into Genomer to allow
annotation coordinates to be automatically updated following such changes to
the sequence.

@@

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs in the
  scaffold. Any unresolved gap regions can also be specified in this file.
  Paired-end sequencing or comparison to reference genomes are a source of
  contig orientation and order. In our genome projects we determine contig
  order by aligning contigs to a reference genome using nucmer
  \cite{kurtz2004}.}

  \item{The \verb+genomer view+ command can be used to generate files of the
  assembled scaffold. These may include the FASTA file of the assembly or a
  GFF3 file of the annotations in the assembled genome. These files can then be
  used for downstream analysis of the sequence.}

  \item{The genome scaffold can be continuously updated through closing any
  remaining gaps. During our microbe sequencing projects we closed gaps through
  a combination of Sanger PCR and further detailed \emph{in silico} analysis.
  Any changes to the scaffold file automatically propagate to generated files
  when the \verb+genomer view+ command is rerun.}

  \item{When researchers are satisfied with the status of their draft the
  required files for submission to GenBank can be generated: assembled FASTA,
  contig FASTA, annotation tables, and scaffold layout in AGP. Additional
  options are used to add the required prefix to gene annotation IDs and begin
  locus ID numbering at the origin of the sequence. Genomer does not produce
  the .asn file which must be submitted to GenBank, but instead the files for
  the input to tbl2asn or sequin. }

>>>>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. In contrast there remains a
steep learning curve in the subsequent computational steps required to process
raw sequence reads into a database-deposited genome sequence. Here we describe
``Genomer,'' a tool to simplify the manual tasks of finishing and uploading a
genome sequence to a database. Genomer can format a genome scaffold into the
common files required for submission to GenBank. Genomer simplifies updating a
genome scaffold by allowing a human-readable YAML format file to be edited
instead of large sequence files. Genomer is written as a command line tool and
is an effort to make the manual, and increasingly common, process of genome
scaffolding more robust and reproducible. Extensive documentation and video
tutorials are available at http://next.gs.

@@

Manually editing a genome sequence or annotation is a non-trivial problem
requiring effort on a researcher's part. The large size of FASTA sequence files
or general feature format (GFF) annotation files makes changing these in a text
editor difficult. Tasks such as organising project files, adding additional
sequences to an assembly, updating incorrect annotations or preparing files for
database deposition all divert effort from the main goal of analysing a genome.
As a result genome projects have a steep learning curve for researchers
embarking on genome sequencing for the first time, while still requiring effort
for experienced researchers.

@@

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer'', to ease and simplify many of the
common steps in our genome project. Several graphical user interface tools are
available to work with genome projects \cite{tanaka2006, wilkinson2002,
lopez2011, carver2012, gordon2003}. Genomer instead provides a simple, command
line interface for genome finishing allowing scriptable automation and
reproducible results. We have open-sourced this tool under the MIT license and
here we describe its application in a microbial genome project.

@@

  \item[Simple editing of a draft genome sequence.]{Genomer is built on the
  existing Scaffolder \cite{barton2012} file format for assembling draft genome
  sequences. This is format requires only the order and IDs of each contig be
  specified. This thereby simplifies the process of re-organising and trimming
  contigs in draft genome without having to edit large nucleotide sequences
  manually.}

  \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
  coordinates of contig annotations to respective positions in the draft genome
  sequence. This allows the scaffold file to be continuously improved and
  updated even after annotation. For instance, additional PCR sequences can be
  used to close gaps in the assembly without reannotation or manually updating
  annotation coordinates.}

  \item[Generation of files for submission to GenBank.]{Submitting a genome
  sequence to GenBank database requires generating specific files. These may
  include a FASTA file of the draft sequence, a table of annotations, a FASTA
  file containing the individual contig sequences, and `a golden path' (AGP)
  \cite{agp-spec} file describing the placement of contigs. Genomer automates
  the generation of all these files from the scaffold file and corresponding
  GFF3 file.}

@@

Genomer is written for the command-line to use in shells such as bash or zsh.
Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. This is an advantage of command-line tools
versus GUI-based tools, the latter are harder to automate and share steps.
Therefore Genomer may be of particular interest to bioinformaticians who prefer
automated approaches and scripting on the command line.

Additionally as Genomer uses plain text it is easy to store versions of the
project using a revision control system such as git \cite{git-scm}. This allows
tracking changes to the project and reverting errors back to earlier versions.
Example build files from our genome projects can be found on GitHub for a
simple plasmid \cite{plasmid-github} and a more complex genome
\cite{genome-github} sequence.

We have released Genomer as open source on GitHub \cite{genomer-github} and
have prepared documentation and video tutorials at http://next.gs. As we have
used this tool extensively to simplify our own genome projects we believe
genomer may also be useful to others in the field. We will illustrate the
possible application of genomer with an example use case.

\subsection*{Genomer use case}

@@

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs in the
  scaffold. Any unresolved gap regions can also be specified in this file.
  Paired-end sequencing or comparison to reference genomes are a source of
  contig orientation and order. In our genome projects we determine contig
  order by aligning contigs to a reference genome using nucmer
  \cite{kurtz2004}. A detailed overview of the scaffolder file format can be
  found on the genomer website.}

  \item{The \verb+genomer view+ command can be used to generate files of the
  assembled scaffold. These may include the FASTA file of the assembly or a
  GFF3 file of updated annotations corresponding to the assembled genome
  sequence. These files can then be used for downstream analyses of the
  sequence.}

  \item{The genome scaffold can be continuously updated through closing any
  remaining gaps. During our microbe sequencing projects we closed gaps through
  a combination of Sanger PCR and \emph{in silico} analysis. Changes to the
  scaffold file automatically propagate updates in the generated files when the
  \verb+genomer view+ command is rerun.}

  \item{When researchers are satisfied with the status of their draft the
  required files for submission to GenBank can be generated: assembled FASTA,
  contig FASTA, annotation tables, and scaffold layout in AGP. Additional
  options are used to add the required prefix to gene annotation IDs and begin
  locus ID numbering at the origin of the sequence. Genomer does not produce
  the .asn file which must be submitted to GenBank, but instead the files that
  may be used as the input to tbl2asn or sequin. }
